---
title: "Problem Set 1 — Democracy Data: Wide to Long"
subtitle: "POLS 6394 · Due Thursday, September 24 at 10:00 AM"
format:
  html:
    embed-resources: true
---

This problem set extends the democracy-data example from 04b. Complete it without AI assistance. You may use the course materials, package documentation, books, Stack Overflow, your classmates, Amanda, and me.

## Setup

Run this chunk as written.

```{r pset01-setup}
#| message: false
library(tidyverse)

dem_wide <- read_csv("Data/dem_score.csv", show_col_types = FALSE)
dem_wide
```

The source is the democracy-score example adapted from [ModernDive, Chapter 4](https://moderndive.com/v2/04-tidy.html#sec-case-study-tidy). Scores run from -10 to 10, with larger values representing more democratic political systems.

## Task 1: Describe the original table

Before reshaping, answer three questions:

1. What does one row represent?
2. Which column identifies the row?
3. Why do the year columns make this a wide rather than tidy table?

Answer: `___`

## Task 2: Reshape the table

Create `dem_long` with three columns: `country`, `year`, and `democracy_score`. Convert `year` from text to an integer.

```{r pset01-long}
#| eval: false
# Fill in, then delete the eval:false line and re-render.
dem_long <- dem_wide |>
  pivot_longer(
    cols = ___,
    names_to = "year",
    values_to = "democracy_score"
  ) |>
  mutate(year = ___)

dem_long
```

## Task 3: Verify the reshape

Show both checks in code:

1. The long table should contain `nrow(dem_wide) * (ncol(dem_wide) - 1)` rows.
2. Each combination of `country` and `year` should occur exactly once.

Report the expected and observed row counts. The duplicate-key check should return no rows.

```{r pset01-checks}
#| eval: false
# Add both validation checks, then delete the eval:false line and re-render.
```

What did the checks show? `___`

## Task 4: Calculate change from 1952 to 1992

Starting with `dem_long`:

1. Keep 1952 and 1992.
2. Reshape those two years into columns named `score_1952` and `score_1992`.
3. Calculate `change = score_1992 - score_1952`.
4. Display the three largest increases and the three largest decreases.

Add the code below. Preserve enough output to verify the country names and changes.

```{r pset01-change}
#| eval: false
# Add the change calculation and both ranked results, then delete the eval:false line and re-render.
```

Largest increases: `___`

Largest decreases: `___`

## Task 5: Explain the representation

In two or three sentences, explain why the long table is easier to group by year and would be easier to plot than the original wide table.

Answer: `___`

## Submit

1. Render this document successfully.
2. Inspect the rendered output and confirm that both validation checks appear.
3. Commit the completed QMD with a specific message.
4. Push before **10:00 AM Thursday, September 24**.
5. Open your private repository on GitHub and confirm that the commit appears.

If rendering fails, record the exact error here, commit the QMD, and push what you completed: `___`
