---
title: "Workbook 03b — Import and the Pipe"
subtitle: "POLS 6394 · Thursday, September 10"
format:
  html:
    embed-resources: true
---

## Today's goal

Import CSV and Stata data, then use pipes and logical conditions to prepare a table. The final four tasks apply the tools to Chicago Airbnb listings.

## RStudio keyboard shortcuts

| Action | Mac | Windows |
|---|---|---|
| Save | **Cmd+S** | **Ctrl+S** |
| Run the current line or selection | **Cmd+Return** | **Ctrl+Enter** |
| Insert `<-` | **Option+−** | **Alt+−** |
| Insert a pipe | **Cmd+Shift+M** | **Ctrl+Shift+M** |
| Insert an R chunk | **Cmd+Option+I** | **Ctrl+Alt+I** |
| Render | **Cmd+Shift+K** | **Ctrl+Shift+K** |

If the pipe shortcut inserts `%>%`, enable **Use native pipe operator** in RStudio's Code options.

## Setup

Run this setup chunk as written. It loads the tidyverse and reads the Airbnb data for the later lab. We'll read the state-policy file together in the first section.

```{r lab03b-setup}
#| message: false
library(tidyverse)
airbnb <- read_csv("Data/airbnb/airbnb.csv", show_col_types = FALSE)
```


# Importing Data

## The tidyverse

The **tidyverse** is a collection of R packages with a shared approach to working with data.

- Tidy tables: one observation per row, one variable per column.
- Named operations connect importing, transforming, and presenting data.
- We'll mainly use tidyverse code, with occasional **base R** examples.

Base R comes with R itself. Tidyverse packages add to it.

Hadley Wickham is a coauthor of [dplyr](https://dplyr.tidyverse.org/authors.html), whose verbs we'll use to transform tables.

## `read_csv()`: the default CSV reader

A CSV is a plain-text table with columns separated by commas. `read_csv()` from `readr` reads one into R.

```{r lab03b-read-csv}
#| message: false
library(tidyverse)

policy <- read_csv(
  "Data/state_policy/cspp_states.csv",
  show_col_types = FALSE
)
```

`<-` assigns the imported table to `policy`. The path starts inside this project. `show_col_types = FALSE` hides a report we'll examine in a moment.

Insert `<-`: **Mac Option+− · Windows Alt+−** (minus key).
Run the selected code: **Mac Cmd+Return · Windows Ctrl+Enter**.

## New lines and indentation

```r
policy <- read_csv(
  "Data/state_policy/cspp_states.csv",
  show_col_types = FALSE
)
```

- Short calls can stay on one line. Split longer calls after `(`, with one argument per line.
- Indent continued lines **two spaces**. Align the closing `)` with the start of the call.
- Keep commas between arguments. A new line doesn't replace a comma.
- Put spaces around `<-` and `=`. Use a blank line between separate steps.

Indentation is the space at the start of a line. Use spaces rather than literal tab characters; the Tab key can insert spaces when configured that way in RStudio. Match the two-space indentation shown here.

The open `(` tells R the call continues. Indentation makes it easier for us to read.

Reformat selected R code: **Mac Cmd+Shift+A · Windows Ctrl+Shift+A**.

## Vectors

A **vector** is an ordered collection of values. The columns we've been working with are vectors.

```{r lab03b-vectors}
scores <- c(0.2, -0.1, NA)
scores
```

- `c()` combines values; commas separate them.
- `<-` saves this three-element vector as `scores`.
- `NA` marks a missing value. It still occupies a position.
- The printed `[1]` marks the first position, not another value.
- R's functions often operate on a whole vector at once. This is usually simpler and faster than processing its values one at a time.

These simple vectors hold one type of value: numbers, text, or logical values.

## Data frames and tibbles

A **data frame** is a rectangular table: named columns of equal length. Each column is a vector; different columns can hold different types.

A **tibble** is a kind of data frame used by the tidyverse. `read_csv()` returns one.

| Behavior | Base data frame | Tibble |
|---|---|---|
| Printing | Can fill the Console | Compact preview with column types |
| Row labels | Can store special row names | Keep identifiers in a regular column |

Most R tools accept either. `as_tibble(x)` converts to a tibble; `as.data.frame(x)` converts to a base data frame. Here `x` means your table.

In base R, `policy$year` identifies both the data frame and its `year` column. Inside a dplyr verb, the data frame has already been supplied, so we can refer to `year` by itself.

[More about tibbles](https://tibble.tidyverse.org/articles/tibble.html). We'll explain dollar-sign column selection in the base-R comparison later.

## Inspecting a tibble

```{r lab03b-glimpse}
glimpse(policy)
```

`glimpse()` lists the dimensions, column names, types, and a few values from every column. `<chr>` means text; `<dbl>` means numbers stored as doubles.

## Read the table before changing it

- **Observational unit:** one state in one year
- **Identifiers:** `st` and `year` together locate a row
- **Measurements:** `pid`, `ideo`, and `statemin` describe that state-year
- Storage type doesn't determine meaning. A numeric case number still identifies a row.

Write down the unit before calculating anything. It determines what your results describe.

::: {.callout-note}
[Sources]
- https://moderndive.com/v2/01-getting-started.html#sec-exploredataframes
:::

## Import functions

| Source | Package | Function |
|---|---|---|
| CSV | readr | `read_csv()` |
| Excel | readxl | `read_excel()` |
| Stata / SPSS | haven | `read_dta()` / `read_sav()` |
| Google Sheets | googlesheets4 | `read_sheet()` |
| One saved R object | base R | `readRDS()` |

You may see `readr::read_csv()`: `::` selects a function from a specific package without loading it with `library()`. We've loaded readr through `library(tidyverse)`, so `read_csv()` is enough.

After importing, use `glimpse()` to check the table.

## Stata data in R

```{r lab03b-stata}
library(haven)

states_stata <- read_dta(
  "Data/state_policy/state_data_1993_2016.dta"
)
glimpse(states_stata)
```

`library(haven)` loads the Stata reader and its label-handling functions.

`read_dta()` reads Stata's file format into a tibble. Use the same inspection and dplyr verbs as with a CSV.

Use `glimpse()` to check the imported column names and types.

## RDS and `.RData`

```{r lab03b-rds}
#| eval: false
dir.create("Processed/03b", recursive = TRUE, showWarnings = FALSE)
saveRDS(policy, "Processed/03b/policy.rds")
policy_again <- readRDS("Processed/03b/policy.rds")
save(policy, file = "Processed/03b/policy.RData")
load("Processed/03b/policy.RData")
```

`dir.create()` makes this lesson's subfolder. `recursive = TRUE` also creates the parent `Processed` folder if needed; `showWarnings = FALSE` suppresses the warning if it exists.

`saveRDS()` writes one object; `readRDS()` returns it for assignment.

`save()` stores named objects; `load()` restores those names. An `.RData` file may be a deliberate bundle, not a whole workspace. It's fine to use, but `load()` doesn't return a table for a pipeline.

## Parsing during import

R must decide whether each column contains numbers, dates, text, or something else. Here the percent sign makes `turnout` text.

```{r lab03b-parse-read}
turnout_raw <- read_csv(
  "Data/import/turnout-messy.csv",
  na = "N/A"
)

glimpse(turnout_raw)
```

`na = "N/A"` identifies the file's missing-value code. If `read_csv()` reports parsing failures, inspect them with `problems()`.

# The Pipe

## Why use a pipe?

A pipe starts with the table and passes it to a function. Here both forms display the first three rows.

```{r lab03b-pipe}
# Function call
head(policy, 3)

# Piped
policy |>
  head(3)
```

`|>` passes the result on its left into the first argument of the function on its right.

Put each pipeline step on a new line, indented two spaces, with `|>` at the end of the preceding line.

## Two pipes you will encounter

```{r lab03b-two-pipes}
# Base R pipe, built into R 4.1 and later
policy |>
  head(3)

# magrittr pipe, loaded with the tidyverse
policy %>%
  head(3)
```

Both pipelines give the same result. Older code often uses `%>%`; we'll use `|>`.

Insert a pipe: **Mac Cmd+Shift+M · Windows Ctrl+Shift+M**.
If this inserts `%>%`, enable **Use native pipe operator** in RStudio's Code options to get `|>`.

# Logic and the Verbs

## Boolean values

A comparison returns a **Boolean**, also called a logical value: `TRUE`, `FALSE`, or `NA` when R can’t tell.

| Operator | Meaning | Example |
|---|---|---|
| `==`, `!=` | equal, not equal | `year == 2010` |
| `>`, `>=`, `<`, `<=` | numeric comparison | `pid > 0` |
| `&`, `\|` | and, or | `year == 2010 & pid > 0` |
| `!` | not | `!is.na(pid)` |
| `%in%` | matches any value in a vector | `year %in% c(2000, 2010)` |

One `=` names an argument or assigns inside `mutate()`. Two `==` test equality.

`is.na(pid)` tests whether `pid` is missing. The `!` reverses the result, so `!is.na(pid)` means that `pid` is present.

## `filter()`: keep rows

`filter()` evaluates a comparison for every row and keeps the rows where the result is `TRUE`.

```{r lab03b-filter}
policy |>
  filter(year == 2010 & pid > 0) |>
  head(6)
```

`&` requires both comparisons to be true. Use `|` when either comparison may be true, and `%in%` when several exact values are acceptable.

## Missing comparisons

```{r lab03b-missing-comparisons}
scores > 0
!is.na(scores)
```

The first line compares each element of `scores` with zero and returns a logical vector of the same length: `TRUE`, `FALSE`, `NA`.

`is.na()` identifies missing values; `!` reverses the result. `filter()` keeps only `TRUE`, so a missing comparison drops that row too.

## Combining conditions

```{r lab03b-combine-conditions}
policy |>
  filter(year == 2010 & (st == "TX" | st == "CA"))
```

`&` requires both sides. `|` accepts either side. Parentheses keep the two state choices together.

Change `"CA"` to another state abbreviation and run it once. `st %in% c("TX", "CA")` is a shorter way to express the original state choices.

## Your Turn: filter

[*~4 minutes.*]{style="font-size: 0.8em;"}

Keep 2010 observations from either the South or the West (`"south"` and `"west"` in `region.name`). Write the condition using `&`, `|`, and parentheses.

```{r lab03b-yt-filter}
#| eval: false
# Fill in, then delete the eval:false line and re-render.
policy |>
  filter(___)
```

Check that every row meets both requirements: the year and one of the two regions. Could you use `%in%` instead?

## `select()` and `arrange()`

`select()` keeps columns. `arrange()` orders rows.

```{r lab03b-select}
policy |>
  filter(year == 2010) |>
  select(state, pollib_median) |>
  arrange(desc(pollib_median))
```

`arrange()` sorts from low to high. `desc()` reverses the order.

## Your Turn: select and arrange

[*~4 minutes.*]{style="font-size: 0.8em;"}

For those same regions in 2010, show `state` and `statemin` (recorded state minimum wage), highest first. Then add a condition keeping only wages of at least $8.

```{r lab03b-yt-select}
#| eval: false
# Fill in, then delete the eval:false line and re-render.
policy |>
  filter(___) |>
  select(state, ___) |>
  arrange(___)
```

Which entries remain? The data include DC as well as states.

## Base R and tidyverse

The same row-and-column operation can be written either way.

```{r lab03b-base-versus-tidy}
base_result <- policy[
  policy$year == 2010,
  c("state", "pid")
]

tidy_result <- policy |>
  filter(year == 2010) |>
  select(state, pid)
head(base_result)
head(tidy_result)
```

Inside base R's brackets, the row instruction comes before the comma and the column instruction comes after it. `$year` extracts the `year` column. We use dplyr because the named verbs remain readable as pipelines grow.

## `mutate()`: calculations and comparisons

`mutate()` creates or replaces columns. The expression on the right can calculate a number or run a comparison.

```{r lab03b-mutate}
policy |>
  filter(year == 2010) |>
  mutate(
    min_wage_hike = statemin - 7.25,
    moredem = pid > 0
  ) |>
  select(state, pid, moredem, min_wage_hike) |>
  head(6)
```

`moredem` is a Boolean column. The original `policy` object remains unchanged because we didn't assign this result back to it.

## More than two categories with `case_when()`

```{r lab03b-case-when}
policy |>
  filter(year == 2010) |>
  mutate(
    partisan_group = case_when(
      is.na(pid) ~ "Missing",
      pid > 0.20 ~ "Democratic",
      pid > 0 ~ "Leans Democratic",
      .default = "Republican or tied"
    )
  ) |>
  select(state, pid, partisan_group) |>
  head(6)
```

`case_when()` checks from top to bottom and uses the first true condition. `~` separates each condition from its result. `.default` handles everything that has not matched.

# Combining Transformations

## A multi-step pipeline

```{r lab03b-pipeline}
policy |>
  filter(year %in% c(1990, 2010)) |>
  mutate(moredem = pid > 0) |>
  select(state, year, pid, moredem, pollib_median) |>
  arrange(year, desc(pollib_median)) |>
  head(6)
```

The pipeline keeps two years, creates a Boolean column, selects five columns, and sorts the result. Each line changes one part of the table.

`c(1990, 2010)` combines the two years into one vector. `%in%` tests whether each value of `year` appears in that vector.

# Lab

[*~20 minutes.*]{style="font-size: 0.8em;"}



Four people are traveling to Chicago for two nights. Find an entire-home listing they could book for at most $600 total. Treat `price` as the nightly listing price for this exercise, excluding fees and taxes. This is a historical dataset, not a current booking search.

## Task 1: Screen the listings

[*~4 minutes.*]{style="font-size: 0.8em;"}



Run this starting screen. `candidates` saves the result so later tasks can use it.

```{r lab03b-count}
glimpse(airbnb)

candidates <- airbnb |>
  filter(
    room_type == "Entire home/apt" & accommodates >= 4 &
      rating >= 4.5
  )
candidates
```

The screen overlooks one booking requirement. Add a condition using `minimum_stay` so a two-night booking is allowed. Keep the rating requirement and inspect the result.

Explain why a listing with `minimum_stay == 3` fails this screen: ___

## Task 2: Calculate the cost of the trip

[*~5 minutes.*]{style="font-size: 0.8em;"}



Starting with `candidates`, calculate the two-night total and each guest's share, keep trips within $600, and show the five cheapest. Retain a listing identifier, neighborhood, and both cost columns.

The column `rownames` identifies a listing in this file. Divide by the four people actually traveling, not the listing's maximum capacity.

```{r lab03b-expensive}
#| eval: false
# Fill in, then delete the eval:false line and re-render.
priced <- candidates |>
  mutate(trip_total = ___, cost_per_guest = ___)

priced |>
  filter(___) |>
  select(rownames, neighborhood, trip_total, cost_per_guest) |>
  arrange(___) |>
  head(5)
```

Cheapest listing ID and cost per guest: ___

## Task 3: Classify access

[*~6 minutes.*]{style="font-size: 0.8em;"}



Use `priced` to create a Boolean `high_access` that is true when either the walk score or transit score is at least 90. Then use `case_when()` to classify the two-night cost:

- `"Under $200"`
- `"$200 to $399"`
- `"$400 or more"`

Put the lowest cutoff first because `case_when()` uses the first true condition.

```{r lab03b-categories}
#| eval: false
# Fill in, then delete the eval:false line and re-render.
classified <- priced |>
  mutate(
    high_access = ___,
    price_group = case_when(
      ___ ~ "Under $200",
      ___ ~ "$200 to $399",
      .default = "$400 or more"
    )
  )

classified |>
  filter(high_access) |>
  select(rownames, trip_total, walk_score, transit_score,
         high_access, price_group) |>
  head(10)
```

Find one row where `high_access` is true. Check its scores to confirm why.

Answer: ___

## Task 4: Make a recommendation

[*~5 minutes.*]{style="font-size: 0.8em;"}



The group cuts its budget to $450 total and still wants `high_access`. Starting with `classified`, write your own pipeline to show five qualifying options, cheapest first. Keep the listing ID, neighborhood, trip cost, both access scores, and price group.

Add a new R chunk for your pipeline: **Mac Cmd+Option+I · Windows Ctrl+Alt+I**. Recommend one listing by ID and explain its cost and access. You may prefer something other than the cheapest option, but it must meet every requirement.

Recommendation and reason: ___

# If Time Remains

## Extension 1: Google Sheets

[*~12–15 minutes.*]{style="font-size: 0.8em;"}



### Step 1: Sign in

We'll create our own spreadsheet and read it back into R using Jenny Bryan's [googlesheets4](https://googlesheets4.tidyverse.org/authors.html) package.

```{r lab03b-sheet-auth}
#| eval: false
library(googlesheets4)
gs4_auth()
```

`library()` loads the package. `gs4_auth()` asks you to choose a Google account and grant access. Follow the browser and Console prompts.

Your account needs permission to create the spreadsheet. We use authentication for this exercise; `gs4_deauth()` is only useful when reading public sheets without signing in.

### Step 2: Create a spreadsheet from mtcars

`mtcars` is built into R: 32 cars with fuel economy and other measurements. Its car names are row names, so make them a column first.

```{r lab03b-sheet-data}
cars <- rownames_to_column(mtcars, var = "model")
```

```{r lab03b-sheet-create}
#| eval: false
cars_sheet <- gs4_create(
  "POLS 6394 - mtcars",
  sheets = list(Cars = cars)
)
```

`gs4_create()` creates the Google spreadsheet. `list(Cars = cars)` supplies one tab named **Cars** containing our table. `cars_sheet` stores its identifier. Run this creation chunk once; rerunning it creates another spreadsheet.

### Step 3: Open your spreadsheet on the web

```{r lab03b-sheet-browse}
#| eval: false
gs4_browse(cars_sheet)
```

`gs4_browse()` opens the spreadsheet in your browser.

- Find the **Cars** tab and the **model** column.
- Change the first car's `mpg` value to 99, an obvious practice edit.
- Wait for Google Sheets to save the change.
- Copy the spreadsheet URL into your workbook so you can return to it.

Spreadsheet URL: `___`

The spreadsheet belongs to your signed-in Google account. You don't need to make it public.

### Step 4: Read the spreadsheet back into R

```{r lab03b-sheet-read}
#| eval: false
cars_back <- read_sheet(cars_sheet, sheet = "Cars")
head(cars_back, 3)
```

`read_sheet()` downloads the tab into a new R object. `sheet = "Cars"` chooses the tab. `head(..., 3)` displays the first three rows.

Confirm that the edited `mpg` value appears. Restore the original value in Google Sheets and rerun this read chunk. The original `cars` object in R still contains that value.

After restarting R, use the quoted spreadsheet URL in place of `cars_sheet`; you don't need to create the spreadsheet again.

### Step 5: Save a local CSV copy

```{r lab03b-sheet-save}
#| eval: false
write_csv(cars_back, "Data/import/mtcars-from-sheets.csv")
```

`write_csv()` writes the downloaded table to the named file inside this project. Open the CSV from RStudio's Files pane.

You can also download it through Google Sheets: **File → Download → Comma-separated values (.csv)**. That saves the current tab through your browser, usually into Downloads.

Keep the Google calls marked `eval: false`: run them manually during class so rendering won't trigger authentication or create another spreadsheet.


If Google sign-in fails in RStudio Server or Posit Cloud, try `gs4_auth(use_oob = TRUE)`. This requests the browser-based sign-in route for a remote R session. Follow its prompts, then continue with the same spreadsheet.

## Extension 2: Stata value labels

A labelled column can store numbers such as 1 and 2 alongside labels such as “Democrat” and “Republican.”

```{r lab03b-stata-labels}
party_codes <- labelled(
  c(1, 2, 1),
  labels = c(Democrat = 1, Republican = 2)
)
party_codes
as_factor(party_codes)
```

This small example constructs labels for demonstration. `as_factor()` turns value labels into categories. Keep numeric measurements numeric.

## Extension 3: Repair a text column

```{r lab03b-parse-number}
turnout <- turnout_raw |>
  mutate(turnout = parse_number(turnout))

glimpse(turnout)
```

`parse_number()` keeps the numeric part of values such as `"53.4%"`. The result is measured in percentage points. Divide by 100 only when a later calculation needs proportions from 0 to 1.

## Extension 4: Repair an import

[*~8–10 minutes.*]{style="font-size: 0.8em;"}

This teaching file uses `"Missing"` for unavailable turnout, percent signs for turnout, and dollar signs for spending.

```{r lab03b-extra-import}
election_raw <- read_csv(
  "Data/import/election-repair.csv",
  na = "Missing",
  show_col_types = FALSE
)
election_raw
```

Inspect the types. Repair `turnout` and `spending`, then find the state with the highest spending per registered voter.

### Check the repaired values

Write a pipeline that parses the two text columns, calculates `spending_per_voter`, and sorts it highest first. Keep the state and the columns needed to check the calculation.

Explain why the row with missing turnout can still have a spending estimate. Does `parse_number()` convert a percentage to a proportion, or just remove the percent sign?

Add your pipeline in a new R chunk, then explain the missing value and percentage handling below.

Explanation: ___

## Extension 5: Repair a screening rule

[*~8–10 minutes.*]{style="font-size: 0.8em;"}

A survey accepts adults aged 18 or older who choose either online or phone participation. This screen lets an ineligible person through.

```{r lab03b-extra-screen}
applicants <- tibble(
  id = c("A", "B", "C", "D", "E"),
  age = c(17, 25, 40, 16, 30),
  mode = c("phone", "online", "mail", "online", "phone")
)
applicants |>
  filter(age >= 18 & mode == "online" | mode == "phone")
```

`tibble()` builds the table from named columns; each `c()` supplies that column's values. Identify the incorrect inclusion before editing the condition.

### Check the corrected screen

- Add parentheses to repair the rule. Keep both participation modes.
- Check every returned row against the written requirements.
- Rewrite the condition with `%in%`. Does it return the same IDs?
- Explain why sorting the original result would not fix it.

Add a new R chunk with both repaired versions.

Incorrect inclusion and explanation: ___

These extensions are optional.

## What I want to remember

- `___`

## What broke, and why

- `___`

## One question I still have

- `___`

## Finalize 03b

- Save and render the workbook as far as you got. Leave unfinished starter chunks at `eval: false`; put `#| eval: false` in any unfinished new chunk so it doesn't run during rendering.
  Save: **Mac Cmd+S · Windows Ctrl+S**. Render: **Mac Cmd+Shift+K · Windows Ctrl+Shift+K**.
- Inspect the changed files in RStudio's Git pane.
- Commit with a specific message, such as `"03b airbnb pipelines"`.
- Push, then open your repository on GitHub and confirm that the commit appears.

If rendering fails, record the error under **What broke, and why** and ask us for help. Still commit and push your saved QMD before leaving; fix the render afterward.

[Keyboard shortcut reference](https://docs.posit.co/ide/user/ide/reference/shortcuts.html). These assume RStudio's default keybindings.
